---
title: "Fictrac Batch Processing"
output: html_document
author: "James Foster"
---
## Details ---------------------------------------------------------------

> **AUTHOR**	James Foster               2022 01 12


> **MODIFIED**	James Foster            2022 01 28


> **DESCRIPTION** 
Loads all `.dat` files saved by `FicTrac` and organises data into
speed and angle. These are saved as compressed `.csv` files and 
which are then combined into a single `.txt` file.


> **INPUTS** 
A `.dat` csv file with 23 columns specified in : 
https://github.com/rjdmoore/fictrac/blob/master/doc/data_header.txt             
The user should specify analysis details (line 70).



> **OUTPUTS** 
Results table (.csv) saved in the same place as the data.


> **CHANGES** 
    - Load FictracDat_Functions.R
    - Speed up with parallel
    - Speed up with data.table
    - Read all files directly to single tab sep file


> **REFERENCES** 
Moore RJD, Taylor GJ, Paulk AC, Pearson T, van Swinderen B, 
Srinivasan MV (2014). 
FicTrac: a visual method for tracking spherical motion and 
generating fictive animal paths. 
*Journal of neuroscience methods* 225, 106–19. 
https://doi.org/10.1016/j.jneumeth.2014.01.010                                  
https://github.com/rjdmoore/fictrac


> **USAGE**  
Fill out user input (lines 70-75)



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Input variables
Find out everything R needs to know to process the `.dat` files.


### - User input
Note that the size of the averaging window an have a big effect on time-averaged parameters: too short and it is biased by small fluctuations, too large and it  may not reflect structures in the data at shorter temporal scales (see: https://en.wikipedia.org/wiki/Aliasing).


I recommend using both `data.table` & `parallel` to speed up processing.
`data.table` is a tool for rapid reading, writing and calculations with large text datasets, and can reduce time by approx. 10%
`parallel` allows processes to be split between different CPUs to be run separately in parallel.


Adding `parallel` to `data.table` can reduce time by a further 55%
On 20220127 a test on AG Pfeiffer laptop 5 (`wbz2084`: Intel(R) Core(TM) i7-10510U CPU @ 1.80GHz   2.30 GHz) with 28 files took 6min 9sec.
```{r User input}
ball_radius = 25#mm
av_window = 5.0#number of seconds to smooth over for averaging
csv_sep_load = ','#Is the csv comma separated or semicolon separated? For tab sep, use "\t"
use_dt = TRUE # whether or not to use data.table for reading, converting & writing
use_par = TRUE # whether or not to use parallel processing
```
### - Load packages
Load the packages required to run the analysis. If these packages are not already installed on your machine, you will need to install them before running this section.
( `install.packages(...)` or select `Tools>Install packages` in the Rstudio menu)
```{r Load packages, echo=TRUE, message = FALSE, warning = FALSE}
require(circular)#for handing angles
if(use_dt){require(data.table)}#for fast data handling
if(use_par){require(parallel)}#for parallel processing
```
### - Load functions
Load the special functions for processing `FicTrac` output.


These should be in a file called `FictracDat_Functions.R` downloaded in the same folder as this script, from GitHub (which is the first place `R` will look).
```{r Get functions, echo=TRUE}
fun_file = "FictracDat_Functions.R"
fun_path = tryCatch(expr = #Search in the folder containing this script
                      {file.path(dirname(sys.frame(1)$ofile), fun_file)},
                    error = function(e){fun_file}
)
if(!file.exists(fun_path))#If not foud, ask the user
{
  msg = paste0('Please select "',fun_file,'"')
  # ask user to find data
  if( Sys.info()[['sysname']] == 'Windows' ){#choose.files is only available on Windows
    message('\n\n',msg,'\n\n')
    Sys.sleep(0.5)#goes too fast for the user to see the message on some computers
    fun_path  <- choose.files(
      default = file.path(gsub('\\\\', '/', Sys.getenv('USERPROFILE')),#user
                          'Documents'),#For some reason this is not possible in the "root" user
      caption = msg
    )
  }else{
    message('\n\n',msg,'\n\n')
    Sys.sleep(0.5)#goes too fast for the user to see the message on some computers
    fun_path <- file.choose(new=F)
  }
}
#read in relevant functions
source(file = fun_path, 
       encoding = 'UTF-8')
```
## Search for the files
The user should select a folder containing all `.dat` files saved by `FicTrac`.


This folder should have the structure:


  Inside the **selected folder** =
      folder named for the **date** (yyyymmdd)
      
      
  Inside each **date folder** = 
      folder with the **animal's name** (any string)
      
      
  Inside each **animal folder** = 
      folder with the **condition name** (any string; should be the same as the equivalent for other animals)
      
      
  Inside each **condition folder** = 
      `.dat` files saved by `FicTrac` for that condition and animal, on that day.
```{r Find the files, echo=TRUE}
# path_folder = FT_select_folder(file_type = 'Fictrac experiment')
#for knitting, it is not possible to ask the user
path_folder = "C:\\Users\\jaf54iq\\Documents\\AllTestdata\\dat_folder"
path_files = list.files(path = path_folder,
                        pattern = '.dat$',
                        recursive = T
)
```
## Set up the parallel cluster
If the user input indicated that the user would like to use parallel processing, `R` will open a "cluster" of processors to use.
It detects how many processors are available and uses all except one (leaving some power for other processes happening on the computer).
In combination with `data.table`, this utilises most of the computer's available resources.
```{r Set up parallel processing, echo=TRUE}
if(use_par){clt = makeCluster(parallel::detectCores() - 1,type="SOCK")}
```
## Process all files
Open each `.dat` file, perform some additional calculations (convert to real-world units of distance, calculate moving averages across time) and then save as a compressed `.csv` file with headers in the same folder. These files can then be opened for plotting (`R` or `Excel`)
```{r Process ".dat" files, echo=TRUE}
message(length(path_files),' files found:\n', paste0(path_files,'\n'), '\nProcessing files.',
        '\n------------------------------------------')
invisible(
  {
  path_input = file.path(
                          path_folder,
                          path_files)
  path_csv = lapply(
                    X = path_input,
                    FUN = FT_read_write,
                    ball_radius = ball_radius,#mm
                    av_window = av_window,#number of seconds to smooth over for averaging
                    csv_sep_load = csv_sep_load,#Is the csv comma separated or semicolon separated? For tab sep, use "\t"
                    speedup_parallel = use_par, #Use the parallel package to speed up calculations
                    speedup_data.table = FALSE, #Use the data.table package to speed up reading and writing
                    compress_csv = TRUE, #Compress to ".gz" to save space?
                    verbose = FALSE#, #tell the user what's going on?
                    # clust = clt
                    )
  names(path_csv) = path_input
  }
  )
if(all(!is.na(path_csv)))
{
  message('\n','Files processed','\n')
}else
{
  warning(path_input[is.na(path_csv)], '\nnot processed!')
}
```
## Combine all files
Read in each of the saved `.csv` files in each folder, and combine the data into a single table including the name of the original file, the condition, the animal and the date. This is then saved as a `.txt` file with the same headers as the `.csv` file, plus headers for condition, bumblebee name, date and `.dat` file.
```{r Combine across all trials, echo=TRUE}
message('Combining data from all trials.',
        '\n------------------------------------------')
  invisible(
    {
      path_txt = FT_combine_folders(
                  path_folder = path_folder,
                  file_type = '_proc.csv.gz',#N.B. currently only for this type
                  speedup_parallel = TRUE, #Use the parallel package to speed up calculations
                  speedup_data.table = TRUE, #Use the data.table package to speed up reading and writing
                  compress_txt = TRUE, #Compress to ".gz" to save space?
                  verbose = TRUE, #Tell the user what is going on
                  recursive = TRUE,   #Search in sub folders                          
                  clust = clt
                  )
    }
  )
if(all(!is.na(path_txt)))
{
  message('\n','Files combined','\n')
}else
{
  warning('\nCombining failed!')
}
stopCluster(clt)
```